Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs

### 2013-02-20 paper.tex
corrections done

\title[Young protoplanetary discs]{Simulated Observations of Young Protoplanetary Discs}
%%authors arent splitting over 2 lines so rawlings is pushed off the edge
%\author[Tom Douglas, Paola Caselli, Et al.]{Tom Douglas$^{1}$\thanks{E-mail:
%pytd@leeds.ac.uk}, Paola Caselli$^{1}$, Aaron Boley$^{2}$, Richard Durisen$^{3}$, Tom Hartquist$^{1}$, %John Ilee$^{1}$, Jonathon Rawlings$^{4}$ \\
%$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK \\
%$^{2}$Department of Astronomy, University of Florida, 211 Bryant Space Center, PO Box 112055, USA\\
%$^{3}$Department of Astronomy, Indiana University, 727 East 3rd Street, Swain West 319, Bloomington, IN %47405, USA\\
%%$^{4}$Department of Physics & Astronomy, University College London, London WC1E 6BT, UK\\
%}

@@

Simulated observations of gravitationally unstable discs have already been performed to study the continuum emission, measure the structure and investigate possible fragmentation (Cossins et al. 2010; Ruge et al. 2013). Molecular line emission from massive discs has been simulated by Krumholz et al. (2007), assuming local thermodynamic equilibrium (LTE). In this paper, we focus on the self-gravitating discs of Boley \& Durisen (2008), where episodic heating induced by spiral shocks is present. This may be a good representation of the earliest phases of protoplanetary discs and an alternative to the young ``static'' discs studied by, e.g., Visser et al. (2009, 2011). As shown by Ilee et al. (2011; hereafter I2011), the spiral shocks cause desorption of volatiles from dust icy mantles and trigger gas-phase chemical reactions with activation energies, too high to occur at lower temperatures. These processes produce clear chemical signatures of the disc dynamics. With the use of 3D radiative transfer modelling, we perform simulated ALMA observations of the disc studied by I2011 and identify the best tracers of the physical structure of self-gravitating discs. The physical, chemical and radiative transfer models are described in Sect.\,\ref{sec:description_model}. Radiative transfer results are in Sect.\,\ref{sec:model_results}, while ALMA simulated observations are in Sect.\,\ref{sec:alma_predictions}. Discussions and conclusions can be found in Sect.\,\ref{sec:discussion}.

@@

\subsection{Physical and chemical structure} \label{subsec:physical_structure}

@@

The pre-stellar core structure is maintained down to a radius of 80\,AU, within which the model of the young protoplanetary disc is used. The disc structure is derived by the hydrodynamic model of Boley (2007) and Boley (2009). The particular model considered here (as well as in I2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating Keplerian disc (figure \ref{velocity}) featuring prominent spiral arms. H$_2$ number densities in the disc range from 10$^{4}$-10$^{13}\,\rm{cm}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The underlying model is sampled as a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$AU in x, y and z. The gas/dust mass ratio was assumed to be 1/100 throughout the disc and the envelope. The opacities were adopted from Ossenkops and Henning (1994) and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history.\newline

Chemical abundances in the disc were taken from I2011 who followed gas-grain chemical processes during the dynamical evolution of the disc. The abundances of 125 species related by 1334 reactions were calculated through the time evolution of the disc. These abundances were interpolated onto a 51$^3$ grid covering the disc with cells of size 2.2$\times$2.2$\times$0.22$\,$AU.  From I2011 we selected the four species which appear to trace different regions of the disc: the inner 20\,AU (OCS), the inner 40\,AU (H$_2$CO),  the region between $\simeq$40 and 60\,AU (HCO$^+$) and one which traces the entirety of the disc (C$^{17}$O) (see figure \ref{Chemistry}, the abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM, Wilson \& Rood 1994). As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also Caselli et al. 2012). As discussed in the result section \ref{sec:model_results}, approximations have been made based on values measured toward similar objects. The molecular data used is from the LAMDA (Leiden Atomic and Molecular DAtabase) database (Sch\"oier et al. (2005; http://home.strw.leidenuniv.nl/$\sim$moldata/).

@@

Simulations are limited to molecules for which we have abundances in both the disc and envelope, and also have calculated Einstein and collisional coefficients. We focus on the frequency range available with ALMA, with particular attention to band 7, which offers the best trade off between resolution and sensitivity. The results presented in this section are limited to those lines which show detectable emission/absorption which can be used to trace either spiral structure or rotation.\newline

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/casa_cont_337GHz.eps}

 \caption{CASA simulation on the continuum emission of the model at 337GHz, using the same ALMA configuration as used for the molecular lines.}
 \label{continuum}
\end{figure}

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/casa_all_spectra.eps}
 \caption{Spectra for each of the simulated species at the positions indicated by the numbers in figure \ref{mom0_maps}, the red green and blue curves are taken from the locations marked 1, 2 and 3 respectively. In each diagram, the three spectra are taken from positions on the Y=0 axis, one is at X=0 (green) and the other 2 are from x=$\pm$ N (plus blue, minus red) where N is the distance from the centre: 10 AU for OCS, 20 AU for H2CO, 28 AU for C17O, 38 AU for HCO+.}
 \label{spectra}
\end{figure*}

@@

One assumption made in this model is that the gas and dust are in thermal equilibrium in the disc. If they are not and the dust is significantly cooler than the gas then transitions may not show up in absorption. However, the large volume densities are expected to provide an efficient dust-gas coupling.\newline
Outflows could contaminate measurements of rotation curves meaning that species  which commonly trace outflows, such as CO and HCO$^+$ are not going to be good tracers of disc rotation. However species such as OCS and C$^{17}$O which are not commonly seen in outflows (e.g. van der Tak et al. 2003, Stanke et al. 2007, Yildiz et al 2012, Ren et al. 2011) can be used to trace disc rotation.\newline

@@

Figures showing different inclinations 
\begin{figure*}
 \includegraphics[width=198mm]{Figures/sim/imageC17O_3-2_15deg_all.eps}

@@

\begin{figure*}
 \includegraphics[width=198mm]{Figures/sim/imageC17O_3-2_45deg_all.eps}

@@

\begin{figure*}
 \includegraphics[width=198mm]{Figures/sim/imageC17O_3-2_75deg_all.eps}

>>>>

\title[Young protoplanetary discs]{Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs}
%%authors arent splitting over 2 lines so rawlings is pushed off the edge
%\author[Tom Douglas, Paola Caselli, Et al.]{Tom Douglas$^{1}$\thanks{E-mail:
%pytd@leeds.ac.uk}, Paola Caselli$^{1}$, Aaron Boley$^{2}$, Richard Durisen$^{3}$, Tom Hartquist$^{1}$, %John Ilee$^{1}$, Jonathon Rawlings$^{4}$ \\
%$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK \\
%$^{2}$Department of Astronomy, University of Florida, 211 Bryant Space Center, PO Box 112055, USA\\
%$^{3}$Department of Astronomy, Indiana University, 727 East 3rd Street, Swain West 319, Bloomington, IN %47405, USA\\
%%$^{4}$Department of Physics & Astronomy, University College London, London WC1E 6BT, UK\\
%}

@@

Simulated observations of gravitationally unstable discs have already been performed to study the continuum emission, measure the structure and investigate possible fragmentation (Cossins et al. 2010; Ruge et al. 2013). Molecular line emission from massive discs has been simulated by Krumholz et al. (2007), assuming local thermodynamic equilibrium (LTE). In this paper, we focus on the self-gravitating discs of Boley \& Durisen (2008), where episodic heating induced by spiral shocks is present. This may be a good representation of the earliest phases of protoplanetary discs and an alternative to the young ``static'' discs studied by, e.g., Visser et al. (2009, 2011). As shown by Ilee et al. (2011; hereafter I2011), the spiral shocks cause desorption of volatiles from dust icy mantles and trigger gas-phase chemical reactions with activation energies, too high to occur at lower temperatures. These processes produce clear chemical signatures of the disc dynamics. With the use of non-LTE 3D radiative transfer modelling, we perform simulated ALMA observations of the disc studied by I2011 and identify the best tracers of the physical structure of self-gravitating discs. The physical, chemical and radiative transfer models are described in Sect.\,\ref{sec:description_model}. Radiative transfer results are in Sect.\,\ref{sec:model_results}, while ALMA simulated observations are in Sect.\,\ref{sec:alma_predictions}. Discussions and conclusions can be found in Sect.\,\ref{sec:discussion}.

@@

\subsection{Physical structure} \label{subsec:physical_structure}

@@

The pre-stellar core structure is maintained down to a radius of 80\,AU, within which the model of the young protoplanetary disc is used. The disc structure is derived by the hydrodynamic model of Boley (2007) and Boley (2009). The particular model considered here (as well as in I2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating Keplerian disc (figure \ref{velocity}) featuring prominent spiral arms. H$_2$ number densities in the disc range from 10$^{4}$-10$^{13}\,\rm{cm}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The underlying model is sampled as a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$AU in x, y and z. The gas/dust mass ratio was assumed to be 1/100 throughout the disc and the envelope. The dust opacities were adopted from Ossenkopf and Henning (1994) and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history.\newline

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011 who followed gas-grain chemical processes during the dynamical evolution of the disc. The abundances of 125 species related by 1334 reactions were calculated through the time evolution of the disc. These abundances were interpolated onto a 51$^3$ grid covering the disc with cells of size 2.2$\times$2.2$\times$0.22$\,$AU.  From I2011 we selected the four species which appear to trace different regions of the disc: the inner 20\,AU (OCS), the inner 40\,AU (H$_2$CO),  the region between $\simeq$40 and 60\,AU (HCO$^+$) and one which traces the entirety of the disc (C$^{17}$O) (see figure \ref{Chemistry}, the abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM, Wilson \& Rood 1994). As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also Caselli et al. 2012). As discussed in the result section \ref{sec:model_results}, approximations have been made based on values measured toward similar objects. The molecular data used is from the LAMDA (Leiden Atomic and Molecular DAtabase) database (Sch\"oier et al. (2005; http://home.strw.leidenuniv.nl/$\sim$moldata/).

@@

Simulations are limited to molecules for which we have abundances in both the disc and envelope, and also have calculated Einstein and collisional coefficients. In the envelope the HCO$^+$ abundance profile follows the H$_2$O profile of L1544 (Caselli et al. 2012), scaled so the maximum is 1$\times$10$^{-8}$ (because photo-disassoc and freeze out are in similar places, need ref). The H$_2$CO has a step profile abundance with a fractional abundance of 1.5$\times$10$^{-8}$ at radii greater than 0.04$\,$pc and 1.5$\times$10$^{9}$ inside it (Young et al. 2004), OCS has a constant abundance of 1.9$\times$10$^{-9}$ (Ren et al. 2011) and C$^{17}$O uses the CO profile of L1544 of KC2010 divided by 1792. We focus on the frequency range available with ALMA, with particular attention to band 7, which offers the best trade off between resolution and sensitivity. The results presented in this section are limited to those lines which show detectable emission/absorption which can be used to trace either spiral structure or rotation.\newline

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/casa_cont_300GHz_invert.eps}

 \caption{CASA simulation on the continuum emission of the model at 300GHz, using the same ALMA configuration as used for the molecular lines. The size of the beam at 300GHz is shown in the lower left corner.}
 \label{continuum}
\end{figure}

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/casa_all_spectra.eps}
 \caption{Spectra for each of the simulated species at the positions indicated by the numbers in figure \ref{mom0_maps}, the red green and blue curves are taken from the locations marked 1, 2 and 3 respectively. In each diagram, the three spectra are taken from positions on the Y=0 axis, one is at X=0 (green) and the other 2 are from x=$\pm$ N (plus blue, minus red) where N is the distance from the centre: 10 AU for OCS, 20 AU for H$_2$CO, 28 AU for C$^{17}$O, 38 AU for HCO$^+$.}
 \label{spectra}
\end{figure*}

@@

One assumption made in this model is that the gas and dust are in thermal equilibrium in the disc. If they are not and the dust is significantly cooler than the gas then transitions may not show up in absorption. However, the large volume densities are expected to provide an efficient dust-gas coupling.\newline
Outflows could contaminate measurements of rotation curves, meaning that species  which commonly trace outflows, such as CO and HCO$^+$ are not going to be good tracers of disc rotation. However species such as OCS and C$^{17}$O, which are not commonly seen in outflows, can be used to trace disc rotation.\newline

@@

Simulations at inclinations other than 30$^\circ$ were also performed with C$^{17}$O in order to demonstate that these results could be replicated over a range of inclination angles. Figures B1-3 show the C$^{17}$O 3$\rightarrow$2 line in the same phisical model but at 15$^\circ$, 45$^\circ$ and 75$^\circ$ to egde on. At 15$^\circ$ to edge on we can still see some structure in the integrated intensity map but also the structure is visible in the position velocity diagram. At larger incliations it gets more difficult to obtain any kinematic information but the physical structure of the disc becomes clearer.

\begin{figure*}
 \includegraphics[width=164mm]{Figures/sim/imageC17O_3-2_15deg_all.eps}

@@

\begin{figure*}
 \includegraphics[width=164mm]{Figures/sim/imageC17O_3-2_45deg_all.eps}

@@

\begin{figure*}
 \includegraphics[width=164mm]{Figures/sim/imageC17O_3-2_75deg_all.eps}
